\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-07-03 paper/subdocuments/methods.Rnw
minor changes methods.Rnw

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe how the two distributions are gone be transformed into a relative distribution, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $x$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(x)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $x$. Also this function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This allows already to spot distributional differences over the whole scale of $Y$ visually. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as $F(y)$, which represents the cumulative probability at point $x$, which is the probability that a randomly chosen value is less than or equal to $x$. The relative distribution of $Y$ to $Y_{0}$ is then defined as

@@

R is obtained from $Y$ by transforming it by the CDF for $Y_{0}$, $\F_{0}$. $R$ therefore measures the relative rank of $Y$ compared to $Y_{0}$.

@@

% Median and shape differences
What we get through the above transformation of two distributions is the overall relative probability density. But differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skewness and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distributions by any measure of central tendency. Here we choose a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% R based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index plays the same role as the difference in Gini-indices, but it can be decomposed to compare the growth in the upper and lower tails. Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is then the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of $y$. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above and below the median (of $g(r)$).

>>>>

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe how the two distributions are gone be transformed into a relative distribution, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $x$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(x)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $x$. This function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This already allows to spot distributional differences over the whole scale of $x$ visually. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as $F(x)$, which represents the probability that a randomly chosen value is less than or equal to $x$. The relative distribution of $Y$ to $Y_{0}$ is then defined as

@@

R is obtained from $Y$ by transforming it by the CDF for $Y_{0}$, $F_{0}$. $R$ therefore measures the relative rank of $Y$ compared to $Y_{0}$.

@@

% Median and shape differences
What we got through the above transformation of two distributions is the overall relative probability density. But differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skewness and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distributions by any measure of central tendency. Here we choose a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% R based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index developed by \citep{Handock and Morris 1999} adresses this issue, because it is decomposable to distinguish differences in the upper and lower tails.  Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is defined as the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of $y$. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above and below the median (of $g(r)$).
